---
title: "Run GENESPACE vignettes - Cotton genomes treated as tetraploids"
author: "JTLovell"
date: "22-Feb 2022"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GENESPACE)
```

# 1. Global Parameters

```{r globParam}
baseDir <- "/Users/jlovell/Desktop/GENESPACE_data/results/"
rawAnnotationDir <- "/Users/jlovell/Desktop/GENESPACE_data"
mcscanDir <- "/Users/jlovell/Documents/comparative_genomics/programs/MCScanX"
nThreads <- 6
path2of <- "orthofinder" # since orthofinder is in the path via conda
```


# 2. Set parameters

**NOTE**: Wheat, Maize and Switchgrass WGD occurred after MRCA of all genomes below. Therefore, treat them as polyploid.

Since some ploidy > 1, run orthofinder again in syntenic blocks. 

```{r rhoParam}
cotton <- list(
  wd = file.path(baseDir, "cotton4x_outgroup"),
  speciesIDs = "cotton",
  genomes = data.table(do.call("rbind", list(
    c(genome = "Gbarbadense", version = "Gbarbadense", ploidy = 2),
    c("Gdarwinii", "Gdarwinii", 2),
    c("Gtomentosum", "Gtomentosum", 2),
    c("Tcacao", "Tcacao_v2.1", 1)))))

cottonParams <- list(
  pepString = "fa",
  orthofinderInBlk = T,
  blkSize = 10,
  nGaps = 10)
```

# 3 Initialize the rho run with the above specified parameters

```{r initrho}
gparCotton4x <- with(cotton, init_genespace(
  genomeIDs = genomes$genome,
  versionIDs = genomes$version,
  ploidy = genomes$ploidy,
  outgroup = "Tcacao",
  speciesIDs = rep(speciesIDs, length(genomes$genome)),
  orthofinderInBlk = TRUE,
  pepString = "fa",
  wd = wd,
  nCores = nThreads,
  path2orthofinder = path2of,
  path2mcscanx = mcscanDir,
  rawGenomeDir = rawAnnotationDir))
```

# 4 Parse the raw annotations. 

```{r rhoAnnot}
parse_phytozome(gsParam = gparCotton4x)
```

# 5 Get orthofinder results

```{r copyOver}
gparCotton4x <- set_syntenyParams(gsParam = gparCotton4x)
gparCotton4x <- run_orthofinder(gsParam = gparCotton4x)
```

Also set the synteny parameters as default. 

```{r rhoOf}
gparCotton4x <- set_syntenyParams(
  gsParam = gparCotton4x,   
  blkSize = 10,
  nGaps = 10)
```

# 6 Build synteny data

```{r rhoSyn}
gparCotton4x <- find_orthofinderResults(gsParam = gparCotton4x)
gparCotton4x <- synteny(gsParam = gparCotton4x, overwrite = T)
```

# 7 Print session info

```{r sesh}
gparCotton4x_outgroup <- gparCotton4x
save(gparCotton4x, file = file.path(baseDir, "gparCotton4x_outgroup.rda"))
sessionInfo()
```
