---
title: "Run GENESPACE vignettes - Cotton genomes treated as tetraploids"
author: "JTLovell"
date: "3-Feb 2022"
output: pdf_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GENESPACE)
```

# 1. Global Parameters

```{r globParam}
baseDir <- "/Users/jlovell/Desktop/GENESPACE_data/results/"
rawAnnotationDir <- "/Users/jlovell/Desktop/GENESPACE_data"
mcscanDir <- "/Users/jlovell/Documents/comparative_genomics/programs/MCScanX"
nThreads <- 6
path2of <- "orthofinder" # since orthofinder is in the path via conda
```


# 2. Set parameters

**NOTE**: Wheat, Maize and Switchgrass WGD occurred after MRCA of all genomes below. Therefore, treat them as polyploid.

Since some ploidy > 1, run orthofinder again in syntenic blocks. 

```{r cottParam}
cottonSplit <- list(
  wd = file.path(baseDir, "cotton_split"),
  speciesIDs = "cotton",
  genomes = data.table(do.call("rbind", list(
    c(genome = "GbarbadenseA", version = "Gbarbadense", ploidy = 1),
    c("GdarwiniiA", "Gdarwinii", 1),
    c("GtomentosumA", "Gtomentosum", 1),
    c("GbarbadenseD", "Gbarbadense", 1),
    c("GdarwiniiD", "Gdarwinii", 1),
    c("GtomentosumD", "Gtomentosum", 1)))))

cottonParams <- list(
  pepString = "fa",
  orthofinderInBlk = T,
  blkSize = 10,
  nGaps = 10)
```

# 3 Initialize the cotton run with the above specified parameters

```{r initcott}
gparCotton2x <- with(cottonSplit, init_genespace(
  genomeIDs = genomes$genome,
  versionIDs = genomes$version,
  ploidy = genomes$ploidy,
  speciesIDs = rep(speciesIDs, length(genomes$genome)),
  orthofinderInBlk = TRUE,
  pepString = "fa",
  wd = wd,
  nCores = nThreads,
  path2orthofinder = path2of,
  path2mcscanx = mcscanDir,
  rawGenomeDir = rawAnnotationDir))
```

# 4 Parse the raw annotations. 

Parse full annotations

```{r cottAnnot}
parse_phytozome(gsParam = gparCotton2x)
```

Split into subgenomes

```{r spli}
chk <- lapply(names(gparCotton2x$paths$gff[1:3]), function(i){
  g <- subset(fread(gparCotton2x$paths$gff[i]), grepl("^A", chr))
  p <- Biostrings::readAAStringSet(gparCotton2x$paths$peptide[i])[g$id]
  fwrite(g, file = gparCotton2x$paths$gff[i], sep = "\t", quote = F)
  Biostrings::writeXStringSet(p, filepath = gparCotton2x$paths$peptide[i])
})
chk <- lapply(names(gparCotton2x$paths$gff[4:6]), function(i){
  g <- subset(fread(gparCotton2x$paths$gff[i]), grepl("^D", chr))
  p <- Biostrings::readAAStringSet(gparCotton2x$paths$peptide[i])[g$id]
  fwrite(g, file = gparCotton2x$paths$gff[i], sep = "\t", quote = F)
  Biostrings::writeXStringSet(p, filepath = gparCotton2x$paths$peptide[i])
})
```

# 5 Get orthofinder results

```{r copyOver}
gparCotton2x <- set_syntenyParams(gsParam = gparCotton2x)
gparCotton2x <- run_orthofinder(gsParam = gparCotton2x)
```

Also set the synteny parameters as default. 

```{r cottOf}
gparCotton2x <- set_syntenyParams(
  gsParam = gparCotton2x,   
  blkSize = 10,
  nGaps = 10)
```

# 6 Build synteny data

```{r cottSyn}
gparCotton2x <- find_orthofinderResults(gsParam = gparCotton2x)
gparCotton2x <- synteny(gsParam = gparCotton2x, overwrite = T)
```

# 7 Print session info

```{r sesh}
save(gparCotton2x, file = file.path(baseDir, "gparCotton2x.rda"))
sessionInfo()
```
